---
title: "rCAT: Species Conservation Assessment tools in R:"
author: "Justin Moat"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rCAT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Vignette Info

Version 0.1. JM just getting his ideas together

## Index


1. Introduction
- 1.1 brief History of development
2. Setup
3. Obtaining occurrence data 
- 3.1 GBIF
- 3.2 CVS
- 3.3  Example datasets
4. Quick walk-through (just to get you started)
5. Worked example for one species (from data to assessment to IUCN)
6. Routines in detail
- 6.1  Simulated datasets
- 6.2  Data check
- 6.3  Projecting you data
- 6.4  EOO
- 6.5  AOO
- 6.6  Sub-populations/number of locations
- 6.7  Batch processing
- 6.8  Other routines
7. Conclusion
8. Acknowledgments
9. References
  
## 1. Introduction


### 1.1 Brief history

- 2003, first tool produced for ArcView 3.0 as used in Willis et al. 2003
- 2007, public release of cats.avx = Extension for ESRI ArcView 3.1 (Moat. 2007)
- 2011, release of GeoCAT online tools geocat.kew.org (Bachman et al 2011)
- 2017, release of rCAT 1.5
- 2020, minor update to rCAT 1.6
- 2020, major update to rCAT 2.0  

## 2. setup

Major release versions are on cran
```
install.packages('rCAT')

```

To install the development version from github you will need devtools package
```
install.packages('devtools')
devtools::install_github("gistin/rCAT")
#this will fail until it's opened up
```

## 3. Obtaining occurance data

The routines in this package are driven in the main by point occurrence data. This data is generally in latitude and longtitude data. 

### 3.1 GBIF data

One of the major repositories for this data is [GBIF] (https://www.gbif.org/). [rgbif] (https://cran.r-project.org/web/packages/rgbif/index.html) is an excellent package to get access to this data.
```
install.packages('rgbif')

```

### 3.2 CSV data
r can except comma-separted values (csv) files as well as many other text file formats. This format can be download from many resources or readily exported from excell.

```
df <- read.csv
```

### 3.3 Example dataset

We have supplied a number of test datasets, this are for fictional species

#### 3.3.1 Polar species
#### 3.3.2 Pan tropical species
#### 3.3.3. Dateline species
#### 3.3.4.1 African continental species
#### 3.3.4.2 American continental species
#### 3.3.4.2 American continental species
```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

## 4. Quick walkthrough

```{r, message=FALSE,fig.show='hold'}
#Prepare your data
#In this case we'll make some data for Cyprus
library(rCAT)
library(sf)
library(ggplot2)
Cyprus <-   tdwg3[tdwg3$LEVEL3_COD=="CYP",]
pts <- as.data.frame(st_coordinates(st_sample(Cyprus,29)))
#set names to lat long
names(pts) <- c('long','lat')
#simple plot of this
ggplot(data=Cyprus) + geom_sf() + geom_point(data=pts,aes(long,lat))
#project your points into something useful i.e. not lat long
#project and center point are automatically set by the wizard
#you can set the center your self ie propts <- simProjWiz(pts,c(33.25,34),returnV='S')
##TD seems to set lat_ts = 0 it should be the centre or maybe not check original reference, I don't think it matters for equatorial projections, as it's just a shift
thepoints <- simProjWiz(pts,returnV='S')
thepointsSF <- simProjWiz(pts,returnV='SF')
#check project details are stored
attr(thepoints,'crs')
#lets check the EOO, it should be approximately the area of Cyprus's ~ 9,000 km2
eoo(thepoints)
#plot this
eeoploy <- eoo(thepoints,"SF")
cplot <- ggplot(data=Cyprus) + geom_sf() + geom_sf(data=thepointsSF) + geom_sf(data=eeoploy,fill=NA,col='green')
cplot
#EOO rating
ratingEoo(eoo(thepoints))
#get AOO should be ~ 116 km2
aoo(thepoints)
#get AOO as cell polygons
aoopolys <- aoo(thepoints,2000,"SF")
#plot it
eooaooplot <- cplot + geom_sf(data=aoopolys,fill=NA,col='red')
eooaooplot
#get rating from aoo
ratingAoo(aoo(thepoints),FALSE)
```


### 6.1 
```{r, message=FALSE,fig.show='hold'}
library(rCAT)
plot(ptsSquare(100,1000),asp=1)
plot(ptsNormal(100,1000),asp=1)
```


Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))


## Old Stuff

[Old pdf for an introduction to this package](https://github.com/gistin/rCAT1/blob/master/vignettes/rCAT.pdf)
